---
title: "Quick start guide for vacamole"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start guide for vacamole}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Set-up
Vacamole can be installed from GitHub
```{r setup}
#library(vacamole)
```
## Overview
`vacamole` is an R package that features a deterministic age-structured compartmental susceptible-exposed-infectious-recovered (SEIR) model and supporting functions. The model was developed to determine the impacts of different COVID-19 vaccination strategies on disease outcomes; therefore the model is an extended SEIR model to include states for severe disease outcomes and vaccination status.

### Model description
The model is a deterministic compartmental SEIR model. The population is partitioned into 10-year age groups (0-9, 10-19, …, 70-79, 80+). Within each age group we further partition the population into those who are unvaccinated, vaccinated with 1 dose, or vaccinated with 2 doses and then finally into disease states: susceptible (S), infected but not yet infectious (E), infectious (I), hospitalized (H), in intensive care (IC), return to the hospital ward after intensive care (HIC), recovered (R), and dead (D) (Figure S1). We assume that individuals who recover from infection cannot be re-infected. When a person is vaccinated, they first enter a hold state where they are vaccinated, but not yet (fully) protected (Shold1d or Shold2d). After a delay period, they enter the vaccinated and protected state for the dose they have received (Sv1d or Sv2d). We assume that vaccination only affects susceptible individuals.

The model is designed to incorporate a single vaccine product with a 2-dose regimen that 1) reduces susceptibility to infection, 2) reduces risk of hospitalization if a vaccinated individual is infected, and 3) reduces risk of infecting others (transmission) if a vaccinated person is infected. The vaccine is assumed to provide “leaky” protection, which means that the vaccine reduces the probability of transitioning between states for each vaccinated person. For example, a vaccine with 70\% effectiveness against infection after the first dose would reduce the probability of transitioning from Sv1d to Ev1d by 70\% compared to someone who is unvaccinated (moving from the S compartment to the E compartment). However, since there is more than one vaccine product currently licensed for use against COVID-19, we incorporate different vaccine products by taking the daily weighted average of the number of people with each vaccine product (and dose), the corresponding delay to protection of each vaccine product, and the vaccine effectiveness against each outcome. Rate of vaccination for each day, vaccine product, and dose for each age group are model inputs. 

### Data
`vacamole` contains several example data sets in the `data/` directory that can be used to run through the package functions. The example data sets include
1. Daily confirmed cases of COVID-19 from the Dutch national notification database Osiris from 1 January 2020 to 22 June 2021 XX cite XX.
2. The vaccination distribution schedules when only adults are vaccinated, when everyone 12 years and older is vaccinated, and when everyone 5 years and older is vaccinated.
3. Contact matrices from the Pienter Corana study XX cite XX prior to the COVID-19 pandemic and in April 2020, June 2020, September 2020, February 2021, and June 2021.

```{r load_data, echo=FALSE, message=FALSE}
# Load required packages ---------------------------------------
# library(dplyr)
# library(knitr)

```
